---
title: "Marianas Trench eDNA"
author: "Charles Nye, Angela Sremba; Oregon State University (OSU)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, Packages, warning=FALSE, include = FALSE}
rm(list = ls()) # Refresh everything
if (!require("pacman")) install.packages("pacman")
pacman::p_load(pacman, tidyverse, data.table, ggtext, ggpubr, viridis, vegan, knitr, devtools) #Some packages might need some finesse
source_url("https://raw.githubusercontent.com/lrjoshi/FastaTabular/master/fasta_and_tabular.R")
#Ensure all packages load
```

```{r, Data Input, include = FALSE}
#Reads table
reads.raw <- read_table(
  "inputs/conv-feature-table.txt"
) %>%
  as.data.frame()

#Blast output
blast.raw <- read.csv(
  "inputs/blast.csv",
  header = TRUE
)

#Metadata
meta.raw = read.csv(
  "inputs/metadata.csv",
  header = TRUE
)

#Convert fasta into sequence list
FastaToTabular(
  "inputs/dna-sequences.fasta"
  )
seqs.raw <- read.csv(
  "dna_table.csv",
  header = TRUE
) %>%
    rename(ASV_ID = name)

#Clean-up input sequence input
seqs.raw$ASV_ID <- sub(".", "", seqs.raw$ASV_ID)
seqs <- seqs.raw[,-1]
```

```{r, Initial Dataframe Construction, include = FALSE}
# Create reads dataframe, use script kung-fu to split and resplice
reads <- reads.raw
rownames(reads) <- reads$ASV_ID
reads$ASV_ID <- rownames(reads)
reads <- reads %>%
  select(ASV_ID,everything())

#Transform into long format
reads <- reads %>%
  pivot_longer(
    cols = 2:ncol(reads),
    names_to = "sample",
    values_to = "read_abundance"
    ) %>%
  filter(read_abundance > 0)

#Calculate pre-filtering ASV richness
ASV.richness.pre <- length(unique(reads$ASV_ID))
```

```{r, ASV Filtering, echo = FALSE, include = FALSE}
# Filter ASVs by certain criteria
reads2 <- reads %>%
  select(c(ASV_ID, read_abundance))
reads2 <- reads2 %>%
  group_by(ASV_ID) %>%
  summarise(read_abundance = sum(read_abundance)) %>%
  filter(read_abundance > 1) #Remove all singletons

# List of ASVs that pass filtering
passing_ASVs <- as.list(reads2$ASV_ID)

# Filter reads dataframe to only include passing ASVs
reads <- reads %>%
  .[reads$ASV_ID %in% c(passing_ASVs), ]

# Calculate passing ASV richness
ASV.richness.post <- length(unique(reads$ASV_ID))
#Calculate percentage of passing ASVs
ASV.percent.pass <- ASV.richness.post / ASV.richness.pre
```


```{r, Reconciling Triplicates, echo = FALSE, include = FALSE}
# Calculate adjusted read counts based on triplicates
# Append metadata
reads <- reads %>%
  left_join(.,
            meta.raw,
            by = "sample")
# Remove superfluous columns, for now
reads <- reads %>%
  select(c(ASV_ID, sample, read_abundance, replicate, triplicate))

# Calculate adjustments
rep.df <- reads %>%
  select(ASV_ID, replicate, read_abundance, triplicate) %>%
  group_by(ASV_ID, replicate) %>%
  filter(triplicate == "yes") %>%
  select(ASV_ID, replicate, read_abundance) %>%
  summarise(read_abundance = sum(read_abundance)/3)
# Round down
rep.df$read_abundance <- floor(rep.df$read_abundance)
```

```{r, Handle additional replicates, echo = FALSE, include = FALSE}
rep.samples = c(
  "MTeDNA_S15",
  "MTeDNA_S15_2",
  "MTeDNA_S25",
  "MTeDNA_S25_2",
  "MTeDNA_S31",
  "MTeDNA_S31_2"
)
rep.df2 = rep.df %>% filter(
  replicate %in% rep.samples
) %>%
  mutate(
    dup = case_when(
      replicate == "MTeDNA_S15" ~ "MTeDNA_S15",
      replicate == "MTeDNA_S15_2" ~ "MTeDNA_S15",
      replicate == "MTeDNA_S25" ~ "MTeDNA_S25",
      replicate == "MTeDNA_S25_2" ~ "MTeDNA_S25",
      replicate == "MTeDNA_S31" ~ "MTeDNA_S31",
      replicate == "MTeDNA_S31_2" ~ "MTeDNA_S31"      
    )
  )
grp = c("ASV_ID", "dup", "read_abundance")
rep.df2 <- rep.df2 %>%
  group_by(across(all_of(grp))) %>%
  summarise(
    read_abundance = (sum(read_abundance))/2
  ) %>%
  rename(
    replicate = dup
  )
rep.df2$read_abundance = floor(rep.df2$read_abundance)
rep.df <- rep.df %>%
  filter(!(replicate %in% rep.samples))
rep.df <- rbind(rep.df, rep.df2)
```


```{r, Appending Inputs, echo = FALSE, include = FALSE}
#Append sequence data to reads dataframe to create primary dataframe
df <- rep.df %>%
  left_join(.,
            seqs,
            by = "ASV_ID")

#Append BLAST results
blast <- blast.raw[,-1]
#Following Pappalardo et al., 2025
#test.blast = filter(
#  blast,
#  phylum %in% c(
#    "Annelida",
#    "Arthropoda",
#    "Chaetognatha",
#    "Chordata",
#    "Cnidaria",
#    "Ctenophora",
#    "Mollusca",
#    "Nemertea",
#    "Porifera",
#    "Rotifera"
#    )
#  )

blast <- blast %>%
  mutate(
    tax.assignment =
      case_when(
        percent_identity < 97.9 ~ phylum,
        TRUE ~ species
      )
  ) %>%
  filter(kingdom != "Fungi" & class != "Arachnida" & class!= "Insecta")

df <- df %>%
  left_join(
    .,
    blast,
    by = "ASV_ID"
  ) %>%
  filter(bit_score > 250) %>%
  filter_at(vars(c(kingdom, phylum, class, order, family)), all_vars(!is.na(.))) %>%
  filter(kingdom != "Fungi" & class != "Arachnida" & class!= "Insecta") #Remove fungi; there are marine fungi, but they are not well-studied or necessarily of interest

meta <- meta.raw %>%
  select(-sample) %>%
  unique()

df <- df %>%
  left_join(.,
            meta,
            by = "replicate") %>% 
  rename(sample = replicate) %>%
  unique() %>%
  as.data.frame()
meta <- meta %>%
  rename(sample = replicate)

#Calculate ASV richness post-taxonomic filtering
df.ASV.richness <- length(unique(df$ASV_ID))
#Calculate species richness post-taxonomic filtering
df.sp.richness <- length(unique(df$species))
#Calculate taxa assignment richness post-taxonomic filtering
df.taxon.richness <- length(unique(df$tax.assignment))
```

```{r, Statistical Analysis Pt. 1, echo = FALSE, include = FALSE}
#Set analysis taxonomic rank
taxon <- "tax.assignment"
grp <- c("sample", taxon)

#Create separate dataframes
df$read_abundance <- as.integer(df$read_abundance)
df.long = df %>% #Long format for plotting
  filter_at(vars(c(lat, long)), all_vars(!is.na(.)))

df.table <- df %>%
  select(c(sample, read_abundance, taxon)) %>%
  group_by(across(all_of(grp))) %>%
  summarise(read_abundance = sum(read_abundance))
df.wide <- df.table %>% #Wide format to create community data
  pivot_wider(
    names_from = taxon,
    values_from = read_abundance
  )
df.wide <- df.wide %>%
  mutate_at(c(2:ncol(df.wide)), ~replace_na(., 0)) %>%
  as.data.frame()

#Create community datasets
comm <- df.wide[,-1] #Read abundance
rownames(comm) <- df.wide$sample
comm.prop <- decostand(comm, method = "total") #Read abundance proportions
comm.log <- decostand(comm, method = "log") #Log-transformed read abundance
comm.log.prop <- decostand(comm.log, method = "total") #Log-transformed read abundance proportions
comm.pa <- decostand(comm, method = "pa") #Presence/absence
comm.hell <- decostand(comm, method = "hellinger") #Hellinger transformation (sqrt of proportions)
comm.ra = comm

remove(comm)
# Set analysis level #
comm = comm.log

#Create metadata for wide format
comm.meta <- df.wide %>%
  left_join(.,
            meta,
            by = "sample") %>%
  select(sample, triplicate:Comments)
```

```{r, Statistical Analysis Pt. 2, echo = FALSE, include = FALSE}
#Shannon Diversity
SW.out <- diversity(comm.ra, index = "shannon")
SW.table <- data.frame(SW.out)
SW.table$sample <- rownames(SW.table)
SW.table <- SW.table %>%
  rename(Shannon_index = SW.out)

comm.meta <- comm.meta %>%
  left_join(.,
            SW.table,
            by = "sample")

#Permutational analysis of variance
set.seed(12031996)
prop.adonis <- adonis2(comm ~ depth_cat + month,
                       data = comm.meta,
                       permutations = 999,
                       method = "bray",
                       by = "terms")
prop.adonis

#Ordination/Nonmetric multi-dimensional scaling
set.seed(12031996)
comm.mds <- metaMDS(comm, dist = "bray", k = 5, trymax = 100)
comm.mds$stress

#Environmental loadings
comm.meta$depth <- as.integer(comm.meta$depth)
env <- comm.meta %>%
  select(depth, Shannon_index)
en = envfit(comm.mds, env, permutations = 999, na.rm = TRUE)
en_coord_cont = as.data.frame(scores(en, "vectors")) * ordiArrowMul(en)
en_coord_cat = as.data.frame(scores(en, "factors")) * ordiArrowMul(en)

data_scores <- as.data.frame(scores(comm.mds, "sites"))
sp_scores <- as.data.frame(scores(comm.mds, "species"))
sp_scores$species <- rownames(sp_scores)
scrs <- scores(comm.mds, display = "sites") %>%
  as.data.frame()
scrs$sample <- rownames(scrs)
scrs <- scrs %>%
  as.data.frame() %>%
  left_join(.,
            comm.meta,
            by = "sample")
```

```{r, Plotting, include = FALSE, echo = TRUE}
cent <- aggregate(cbind(NMDS1, NMDS2) ~ depth_cat, data = scrs, FUN = mean)
segs <- merge(scrs, setNames(cent, c("depth_cat","oNMDS1","oNMDS2")),
              by = "depth_cat", sort = FALSE)
NMDS.plot <- ggplot(scrs, aes(x = NMDS1, y = NMDS2, color = depth_cat)) +
  geom_segment(data = segs, mapping = aes(xend = oNMDS1, yend = oNMDS2), alpha = 0.3) +
  geom_point(data = cent, size = 4) + 
  geom_point(size = 2, alpha = 0.3) + coord_fixed() +
  scale_color_viridis(discrete = TRUE, option = "turbo") +
  geom_segment(aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2), 
               data = en_coord_cont, linewidth = 1, alpha = 1, colour = "black") +
  #geom_point(data = en_coord_cat, aes(x = NMDS1, y = NMDS2), 
  #           shape = "diamond", size = 4, alpha = 0.6, colour = "navy") +
  #geom_text(data = en_coord_cat, aes(x = NMDS1, y = NMDS2+0.04), 
  #          label = row.names(en_coord_cat), colour = "navy", fontface = "bold") + 
  geom_text(data = en_coord_cont, aes(x = NMDS1, y = NMDS2), colour = "grey30", 
            fontface = "bold", label = row.names(en_coord_cont)) + 
  theme(panel.background = element_blank(),axis.line = element_line(colour = "black")) +
  ggtitle("eDNA NMDS Ordination") +
  xlab("NMDS 1") + ylab("NMDS 2") +
  labs(subtitle = "Bray-Curtis Dissimilarity on Log-Transformed Reads") +
  labs(color = "Depth Categories")

plot1 = df.long %>%
  group_by(depth_cat, phylum) %>%
  summarise(read_abundance = sum(((read_abundance)))) %>%
  arrange(desc(read_abundance)) %>%
  ggplot(aes(x = as.factor(depth_cat), y = read_abundance, fill = phylum)) +
  geom_bar(stat = "identity", colour = "black", position = "fill") +
  scale_fill_viridis(discrete = TRUE, option = "turbo") +
  #facet_grid(vars(depth_cat)) +
  theme(panel.background = element_blank(),axis.line = element_line(colour = "black")) +
  coord_flip() + scale_x_discrete(limits = rev) +
  xlab("") +
  ylab("eDNA Read Abundance (RA proportions)") +
  labs(fill = "eDNA Detected Taxa") + ggtitle("A")

plot2 = df.long %>% 
  group_by(depth_cat, phylum) %>%
  summarise(read_abundance = sum((log1p(read_abundance)))) %>%
  arrange(desc(read_abundance)) %>%
  ggplot(aes(x = as.factor(depth_cat), y = read_abundance, fill = phylum)) +
  geom_bar(stat = "identity", colour = "black", position = "fill") +
  scale_fill_viridis(discrete = TRUE, option = "turbo") +
  #facet_grid(vars(depth_cat)) +
  theme(panel.background = element_blank(),axis.line = element_line(colour = "black")) +
  coord_flip() + scale_x_discrete(limits = rev) +
  xlab("") +
  ylab("eDNA Log RA (LRA, proportions)") +
  labs(fill = "eDNA Detected Taxa") + ggtitle("B")
```

*Introduction*

Environmental DNA (eDNA) samples were collected at various depths in the western Pacific Ocean, terminating at the bottom of the deep-sea geological feature known as the Marianas Trench. There are associated hydrophone data collected from this same expedition, but this report focuses on the eDNA findings.

*Wet Lab Methods*

DNA extraction and next-generation sequencing (NGS) library preparation were conducted at OSU Hatfield Marine Science Center (HMSC) in Newport, Oregon. eDNA water samples were filtered through a 1uM polycarbonate track etch (PCTE) filter. DNA was extracted using a phenol chloroform-isoamyl approach. We amplified the "barcode of life" gene, cytochrome C oxidase subunit I (COI), in triplicate assays of the n = `r length(unique(meta$sample))` samples, inclusive of controls. PCR assays were pooled, concentrated, and purified using Beckman Coulter AMPure XP beads. We ligated Illumina NGS indices onto all successful amplicons and pooled samples for sequencing. NGS processing was conducted at the OSU Center for Quantitative Life Studies (CQLS) in Corvallis, Oregon.

*Data Analysis Methods*

We used the program wrapper QIIME2 to remove poor-quality reads and chimera sequences. Amplicon sequence variants (ASVs) were assigned taxonomic identities from a BLAST query of the NCBI Nucleotide database. All data analysis was performed in R. We removed poor quality, off-target, or especially rare ASVs from the dataset. All data analyses used log-transformed proportions of read count/abundance. We used a permutational analysis of variance (PERMANOVA) to quantify compositional differences by depth category. We visualized these differences using a nonmetric multi-dimensional scaling (NMDS) ordination plot.

*Results*

The unfiltered dataset consists of `r ASV.richness.pre` ASVs. After our strict quality parameters, we are left with `r df.ASV.richness` ASVs that further collapse into `r df.sp.richness` species. This value changes to `r df.taxon.richness` when incorporating non-species ASVs collapsed into phyla. The average Shannon diversity in the entire dataset is `r mean(SW.out)`.

```{r, echo = FALSE}
NMDS.plot
```

We find that depth category is a significant (p = `r prop.adonis[1,5]`) predictor of community composition in this system, further suggested by the notable influence of the continuous depth variable in the ordination plot. Furthermore, depth category has a model fit of `r prop.adonis[1,3]`, which is substantial as a predictor but we clearly cannot explain the full variation from this simple model, alone.


```{r, include = FALSE, echo = FALSE}
df.table <- df.table[,-1] %>%
  group_by(tax.assignment) %>%
  summarise(read_abundance = sum(read_abundance)) %>%
  arrange(desc(read_abundance)) %>%
  slice(1:15)
```

`r kable(df.table, col.names = c("Taxonomic Assignment", "Read Count"), "simple", caption = "Table 1: The top 15 most-observed taxa from the Marianas Trench eDNA dataset.")`

The most biodiverse depth category in this region is, perhaps intuitively, the surface or epipelagic. Read detections and biodiversity pick back up towards the hadalpelagic, and the only notable detections near the benthos of this system belong to gastropods. The closest species match is *Atlanta frontieri*, at 98% confidence from GenBank.


```{r, echo = FALSE}
simper.general <- summary(simper(comm, permutations = 999), ordered = TRUE)
simper.depth <- summary(with(comm.meta, simper(comm, depth_cat, permutations = 999)))

simper.depth.table = simper.depth$`01. Epipelagic (0-200m)_04. Benthos (10km)` %>%
  as.data.frame() %>%
  mutate(
    species = rownames(.)
  ) %>%
  select(
    c(
      species,
      average,
      p
    )
  ) %>%
  head(
    n = 15
  ) %>%
  as_tibble() %>%
  tibble::remove_rownames() %>%
kable(
  "simple",
  col.names = c(
    "Species",
    "Avg. Dissim. Contribution",
    "Permuted P-Value"
  )
)

```

`r simper.depth.table`

Using SIMPER analysis (Clarke 1993) between the two depth extremes of surface/epipelagic and the 10km benthos, we see *Atlanta* 

```{r, echo = FALSE, include = FALSE}
deep.df = df %>% filter(depth_cat != "01. Epipelagic (0-200m)")
plot3 = df %>%
  group_by(tax.assignment, depth_cat) %>%
  summarise(read_abundance = sum((read_abundance))) %>%
  ggplot(
    aes(
      x = reorder(tax.assignment, -(read_abundance), sum),
      y = (read_abundance),
      fill = depth_cat
    )
  ) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_viridis(discrete = TRUE, option = "plasma") +
  coord_flip() + scale_x_discrete(limits = rev) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("eDNA Taxa") +
  ylab("RA") + labs(fill = "Depth Category") + ggtitle(
    "Marianas Trench eDNA Taxa Abundance by Depth"
  )
```

```{r, echo = FALSE}
eDNA_barplots = ggarrange(
  plot1, plot2, common.legend = TRUE, legend = "right", nrow = 2
  )
annotate_figure(eDNA_barplots, top = text_grob("mtCOI eDNA Phyla by Depth", 
               color = "black", size = 14))
plot3
```


```{r, echo = FALSE}
depth.A = df %>%
  group_by(tax.assignment, depth_cat) %>%
  summarise(read_abundance = sum(log1p(read_abundance))) %>%
  ggplot(
    aes(
      x = reorder(tax.assignment, -(read_abundance), sum),
      y = (read_abundance),
      fill = depth_cat
    )
  ) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_viridis(discrete = TRUE, option = "plasma") +
  coord_flip() + scale_x_discrete(limits = rev) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("") +
  ylab("LRA") + labs(fill = "Depth Category") +
  ggtitle(
    "A. LRA by Taxonomic Units"
  )

depth.B = df %>%
  filter(type == "eDNA") %>%
  group_by(type, depth_cat) %>%
  summarise(read_abundance = sum(log1p(read_abundance))) %>%
  ggplot(
    aes(
      x = type,
      y = (read_abundance),
      fill = depth_cat
    )
  ) +
  geom_bar(stat = "identity", colour = "black", position = "fill") +
  scale_fill_viridis(discrete = TRUE, option = "plasma") +
  #coord_flip() + scale_x_discrete(limits = rev) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("") + xlab(NULL) +
  ylab("LRA (proportions)") + labs(fill = "Depth Category") + ggtitle(
    "B. All Samples, Aggregated"
  ) + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

depth.all = ggarrange(
  depth.A,
  depth.B,
  common.legend = TRUE,
  legend = "right"
) + bgcolor("white")
annotated.depth.all = annotate_figure(
  depth.all,
  top = text_grob(
    "eDNA Log-Transformed Read Abundance (LRA)", color = "black", size = 16
  )
)
annotated.depth.all

```

and *Acartia* are responsible for the bulk of Bray-Curtis dissimilarity.

```{r, echo = FALSE, include = FALSE}
# Redefine df.wide and df.long #
remove(df.wide)
remove(df.long)
remove(comm)

comm = comm.ra
comm$sample = rownames(comm.ra)
df.wide = comm %>% left_join(
  comm.meta,
  by = "sample"
) %>%
  select(sample, triplicate:Shannon_index, everything())
df.long = df.wide %>%
  pivot_longer(
    cols = 18:ncol(df.wide),
    names_to = "tax.assignment",
    values_to = "comm.index"
  )
```

```{r, echo = FALSE, include = FALSE}
df.long = df.long %>% 
  filter_at(vars(c(tax.assignment, comm.index,
                   depth_cat)), all_vars(!is.na(.)))

grp = c("tax.assignment", "depth_cat")
df.long %>%
  group_by(across(all_of(grp))) %>%
  summarise(comm.index = sum((comm.index))) %>%
  filter(comm.index > 0) %>%
  write.csv(
    "taxa_by_depth.csv"
  )


```


```{r, echo = FALSE, include = FALSE}
eDNA.model = glm(
  comm.index ~ depth_cat + Shannon_index + tax.assignment, family = binomial(link = "log"), data = df.long
)
summary(eDNA.model)
with(summary(eDNA.model), 1 - deviance/null.deviance)

```








